---
title: "Lesson_28_Boardsheet"
author: "Kevin Cummiskey"
date: "4/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 3,
                      message = FALSE,
                      warning = FALSE)
```

## Review

What characteristics of a hitter does OPS quantify?

\vspace{1in}

Why might one use OPS+ or adjusted OPS+ instead of OPS?

\vspace{1in}

What are some limitations of these statistics?

\vspace{1in}

## Linear Weights

Recall we talked about the relationship between Runs and Wins in Chapter 4 of the Marchi text.  What was the rule of thumb we used?

\vspace{1in}

(*Understanding Sabermetrics* pg 83) Ideally, we want a statistic for batters that:

* quantifies their contribution to runs scored.

* is not based upon the situations the batters faced when they came to the plate (since their batting actions did not create those situations).

How about RBI's? A batter is awarded Runs Batted In (RBIs) for most situations when his plate appearance results in runs scored.  Why are RBIs not a good measure of a player's contribution to runs scored?

\vspace{1in}

Various researchers have proposed linear models to quantify a player's contribution to runs scored.  These models weight individual statistics.  For the purposes of today's lesson, let's investigate the condensed model proposed by Thorn and Palmer (*Understanding Sabermetrics*, pg 87). I'll refer to this model as the Linear Weights Model

\[Batting Runs = w_1\text{1B} + w_2\text{2B} + w_3\text{3B} + w_4\text{HR} + w_5 (\text{BB}+\text{HBP}) - w_6(AB-H)\]

How can we get weights for the Linear Weights Model?

\vspace{2in}

How do we interpret $Batting Runs$?

\vspace{1in}

Let's see what this looks like for the 2018 season.

```{r}
library(Lahman)
library(tidyverse)
library(plotly)
library(knitr)

# Calculate weights from run values using Retrosheet play-by-play data
# this code will only run if you have a 2018 retrosheet
# event-by-event data on your computer.
source("../MA388_Solutions/linear_weights.R")
weights <- linear_weights(2018) %>% pluck("weights")
weights %>% kable(digits = 3)
```

\newpage

How do these weights compare to those in slugging percentage (SLG)?

\vspace{2in}


```{r}
# 2018 Players with at least 500 at bats
vars = c("AB", "H", "X2B", "X3B", "HR", "BB",
         "HBP", "SF","RBI")
Batting %>% 
  filter(yearID == 2018) %>% 
  group_by(playerID) %>% 
  summarise_at(vars, sum) %>% 
  filter(AB >= 500) %>% 
  left_join(Master %>% select(nameLast,nameFirst,playerID)) %>% 
  mutate(name = paste(nameFirst, nameLast, sep = " ")) %>% 
  select(name, everything(), -nameLast,-nameFirst) %>% 
  mutate(X1B = H - X2B - X3B - HR,
         SLG = (X1B + 2*X2B + 3*X3B + 4*HR)/AB,
         OBP = (H + HBP + BB)/(AB + HBP + SF + BB),
         OPS = SLG + OBP,
         AVG = H/AB) -> batting.2018

#calculates Batting Runs using Thorn and Palmer's condensed Linear Weights model. 
#note statistics and weights have to be in the same order
batting_runs <- function(statistics, weights){
  runs <- round(sum(statistics*weights),1)
  return(runs)
}

batting.2018 %>%
  group_by(playerID) %>% 
  mutate(batting.runs = batting_runs(statistics = c(X1B, X2B, X3B, BB + HBP, HR, AB - H),
                                     weights = weights %>% pull(weight))) %>% 
  arrange(-batting.runs) %>% 
  group_by()-> batting.2018
  
batting.2018 %>% 
  select(name, AB, H, HR, RBI, AVG, OPS, batting.runs) %>% 
  head(10) %>% 
  kable(digits = 3, 
        caption = "Top 10 MLB Players (with at least 500 at bats) - Batting Runs 2018")
```

How many wins would you attribute to Mookie Betts in the 2018 season?

\vspace{0.5in}

```{r}
batting.2018 %>% 
  select(name, AB, H, HR, RBI, AVG, OPS, batting.runs) %>% 
  tail(10) %>% 
  kable(digits = 3,
        caption = "Bottom 10 MLB Players (with at least 500 at bats) - Batting Runs 2018")
```

Let's see how $Batting Runs$ compare to traditional statistics.

```{r, fig.height=4}
library(gridExtra)

title.size = 12

p.rbi <- batting.2018 %>% 
  ggplot(aes(label = name,
             x = batting.runs,
             y = RBI)) +
  geom_point() +
  labs(title = "Linear Weights vs. RBI") +
  theme_classic() +
  theme(plot.title = element_text(size = title.size))

p.avg <- batting.2018 %>% 
  ggplot(aes(label = name,
             x = batting.runs,
             y = AVG)) +
  geom_point() +
  labs(title = "Linear Weights vs. Batting Average") +
  theme_classic() +
  theme(plot.title = element_text(size = title.size))


p.slg <- batting.2018 %>% 
  ggplot(aes(label = name,
             x = batting.runs,
             y = SLG)) +
  geom_point() +
  labs(title = "Linear Weights vs. Slugging") +
  theme_classic() +
  theme(plot.title = element_text(size = title.size))

p.ops <- batting.2018 %>% 
  ggplot(aes(label = name,
             x = batting.runs,
             y = OPS)) +
  geom_point() +
  labs(title = "Linear Weights vs. OPS") + theme_classic() +
  theme(plot.title = element_text(size = title.size))

grid.arrange(p.rbi,p.avg, p.slg, p.ops, ncol = 2)

ggplotly(p.rbi)
ggplotly(p.avg)
ggplotly(p.slg)
ggplotly(p.ops)
```

Let's say you're a general manager.  Which statistic would you use?

\vspace{1in}

What other factors would you want to consider?

\vspace{2in}

In terms of $BattingRuns$, is it more important to hit for power or on base percentage?

\vspace{1in}

```{r, fig.height=3}
library(viridis)
batting.2018 %>% 
  mutate(OBP.scaled = scale(OBP),
         SLG.scaled = scale(SLG),
         br.scaled = scale(batting.runs)) %>% 
  ggplot(aes(x = OBP.scaled, y = SLG.scaled, color = br.scaled)) +
  geom_point() +
  scale_color_viridis() +
  geom_abline(slope = 1, intercept = 0)
```


```{r, include=FALSE, eval = FALSE} 
# 2018 players
Batting %>% 
  filter(yearID == 2016) %>% 
  group_by(playerID) %>% 
  summarize(X2B = sum(X2B),
            X3B = sum(X3B),
            HR = sum(HR),
            AB = sum(AB),
            BB = sum(BB),
            H = sum(H)) %>% 
  mutate(X1B = H - X2B - X3B - HR) %>% 
  filter(AB > 100) -> batting.2016

Salaries %>% 
  filter(yearID == 2016) %>% 
  group_by(playerID) %>% 
  summarize(salary = sum(salary)) %>% 
  mutate(salary.million = round(salary/1000000,2)) -> salary.2016

batting.2016 %>% 
  left_join(salary.2016, by = "playerID") %>% 
  left_join(select(Master, nameLast, nameFirst,playerID, debut)) %>% 
  mutate(name = paste(nameFirst, nameLast, sep = " ")) %>% 
  select(name,everything(),-nameLast,-nameFirst) -> batting.2016

p <- batting.2016 %>% 
  pivot_longer(cols = X2B:X1B,
               names_to = "statistic",
               values_to = "value") %>%
  filter(statistic %in% c("H", "HR")) %>% 
  ggplot(aes(x = value, 
             y = salary.million,
             label = name)) +
  geom_point() +
  facet_wrap(~ statistic, ncol = 2, scales = "free_x") +
  theme_bw() 

ggplotly(p)
```



```{r, include = FALSE, eval = FALSE}
library(GGally)

plot_coord <- ggparcoord(batting.2016,
           columns = 3:9,
           showPoints = TRUE) +
  
ggplotly(plot_coord)

```