---
title: "Lesson_29_Boardsheet"
author: "Kevin Cummiskey"
date: "4/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.height = 3, message= FALSE, warning = FALSE)
```

## Review

Last class, we discussed Linear Weights models.  Linear Weights models attempt to estimate $BattingRuns$ for each player using individual performance. What are $BattingRuns$?

\vspace{1in}

How do we obtain weights in the model?

\vspace{1in}

What are some limitations of the Linear Weights models?

\vspace{1in}

## Runs Created

In the late 1970s, Bill James developed a series of statistics called *Runs Created (RC)*. These statistics have the common form:

\[RC = \frac{A \times B}{C}\]

where $A$ quantifies how often the player/team gets on base, $B$ quantifies how they advance on base, and $C$ represents the opportunities.  The simplest version of RC is:

\[RC_\text{basic} = \frac{(H + BB)\times TB}{AB + BB}\]

Let's see how well $RC_\text{basic}$ predicts actual runs scored for teams in the 2018 season.

```{r}
library(Lahman)
library(tidyverse)
library(ggrepel)
library(plotly)
library(knitr)

Teams %>% 
  filter(yearID == 2018) %>% 
  mutate(X1B = H - X2B - X3B - HR,
         TB = X1B + 2*X2B + 3*X3B + 4*HR,
         RCbasic = ((H + BB)*TB)/(AB + BB)) -> teams.2018

teams.2018 %>% 
  ggplot(aes(x = RCbasic, y = R, label = teamID)) +
  geom_point() +
  geom_text_repel()
```

Is $RC_\text{basic}$ a *good* model for runs?  What makes it a good model?

\vspace{1in}

A nice feature of $RC$ is a team's $RC$ is very close to the sum of its individual players.  Therefore, we can use $RC$ to assign credit to players for a certain proportion of their team's runs.

```{r}
Batting %>% 
  filter(yearID == 2018) %>% 
  mutate(X1B = H - X2B - X3B - HR,
         TB = X1B + 2*X2B + 3*X3B + 4*HR,
         RCbasic = ((H + BB)*TB)/(AB + BB)) %>% 
  left_join(select(Master,playerID, nameLast, nameFirst)) %>% 
  mutate(name = paste(nameFirst, nameLast, sep = " ")) -> batting.2018.all

batting.2018.all %>% 
  group_by(teamID) %>% 
  summarize(RCbasic.summed = sum(RCbasic, na.rm = TRUE)) %>% 
  right_join(teams.2018) -> teams.2018
  
teams.2018 %>% 
  ggplot(aes(x = RCbasic, y = RCbasic.summed)) +
  geom_point() +
  labs(x = "RC from team statistics",
       y = "RC summed over individual players",
       title = "Team RC and RC summed over players is very close")

batting.2018.all %>% 
  left_join(select(teams.2018, teamID, RCbasic.summed)) %>% 
  group_by(teamID) %>% 
  mutate(RC.team.perc = RCbasic/RCbasic.summed) %>% 
  arrange(-RC.team.perc) -> batting.2018.all


batting.2018.all %>% 
  select(name, teamID, R, RC.team.perc) %>% 
  head(10) %>% 
  kable(digits = 2,
        caption = "Highest Percentage of Team's Runs Scored - 2018")
```

Let's see how $RunsCreated$ compares to $BattingRuns$ from the Linear Weights model.

```{r}
read_csv(file = "data/batting_2018.csv") -> batting.2018

#add runs created to data frame from last class
batting.2018.all %>% 
  group_by(playerID) %>% 
  summarize(RCbasic = sum(RCbasic, na.rm = TRUE)) %>% 
  right_join(batting.2018) -> batting.2018

p1 <- batting.2018 %>% 
  ggplot(aes(x = batting.runs, 
             y = RCbasic,
             label = name)) +
  geom_point() +
  labs(x = "Batting Runs (Linear Weights Model",
       y = "Runs Created - Basic",
       title = "Linear Weights vs. Runs Created")
p1
#ggplotly(p1)
```

Why might we prefer one model to the other?

\vspace{1in}

How can we improve upon the basic version of Runs Created?

\vspace{1in}
