---
title: "Lesson_39_Boardsheet"
author: "Kevin Cummiskey"
date: "5/6/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 3)
```

```{r}
library(tidyverse)
library(baseballr)
library(viridis)
library(lubridate)
```

## Launch Angle/Speed and Hit Probability

Today, we are going to look at how the probability of a hit changes with launch angle and speed.  Much of the code for today's lesson comes from Chapter 12 of *Analyzing Baseball Data with R* by Marchi et al.

https://www.mlb.com/news/new-statcast-metric-barrels-has-best-hit-balls-c201699298

First, let's download StatCast data for 2017 using the scrape\_statcast\_savant function. Darn, it only lets us download 40,000 pitches (about 15 days) at a time.  Ok, maybe I should start doing this:

```{r, eval = FALSE}
#Note R stores dates as the number of days since Jan 1, 1970



#March 31, 2017 to April 13, 2017
scrape_statcast_savant(start_date = as_date(17256), end_date = as_date(17270)) 
  #April 14, 2017 to April 28, 2017
scrape_statcast_savant(start_date = as_date(17271), end_date = as_date(17285)) 

# This is going to get tedious going all the way to September
```

What does splitting, applying, combining mean?

\vspace{1in}

How can we use this concept to make it easier to read in this data?

\vspace{1in}

```{r, cache = TRUE}
dates <- tibble(start_date = as_date(seq(17226, 17454, 15)),
                end_date = as_date(start_date + 14))

dates
```

Now, let's apply the scrape\_statcaset\_savant using the rows of \texttt{dates} as the arguments on the function.  

```{r, cache = TRUE}
dates %>% 
  pmap(scrape_statcast_savant) %>% 
  do.call(rbind, .) %>% 
  filter(type == "X") -> batted2017
```

Now, we have a data frame called \texttt{batted2017} containing all batted balls in the 2017 season.

```{r}
# determine if batted ball was a hit
batted2017 %>% 
  mutate(hit_flag = factor(ifelse(events %in% c("single", "double",
                                         "triple", "home_run"),1,0))) -> batted2017

#batted2017 %>% 
#  select(player_name, stand, hc_x, hc_y, launch_angle, launch_speed, if_fielding_alignment) %>% 
#  write_csv(path = "data/teeStatCast.csv")


#Plot each batted ball by whether it's a hit or not
guidelines <- tibble(
  launch_angle = c(10,25,50),
  launch_speed = 40,
  label = c("Ground balls", "Line drives", "Flyballs")
)

ev_plot <- batted2017 %>% 
  sample_n(nrow(.)/2) %>% 
  ggplot(aes(x = launch_speed, 
             y = launch_angle,
             color = hit_flag)) +
  geom_point() +
  scale_color_viridis(discrete = TRUE) +
  geom_hline(data = guidelines,
             aes(yintercept = launch_angle),
             color = "black",
             linetype = 2) +
  geom_text(data = guidelines,
            aes(label = label, x = 0, y = launch_angle - 4),
            color = "black", hjust = "left")

ev_plot
```

Instead of the plot above, why might we want to view hit probability as a smoothed function of launch speed and angle?

\vspace{1in}

Is a linear function of launch speed and velocity appropriate for this data?

\vspace{1in}

Let's fit a generalized additive model for hit probability as a smoothed function of launch speed and angle. Let $Y_i \sim \text{Bernoulli}(\pi_i)$ equal 1 when batted ball $i$ is a hit, and 0 otherwise.  Our model is:

$$\log\left(\frac{\pi_i}{1 - \pi_i}\right) = s(Angle_i, Speed_i)$$

```{r, cache= TRUE}
library(mgcv)

fit <- batted2017 %>% 
  gam(hit_flag ~ s(launch_speed, launch_angle),
      family = "binomial",
      data = .)
```

Let's look at some predictions from the model. First, let's look at batted balls with an speed for 100mph for various launch angles.

```{r}
batted100mph <- tibble(launch_angle = seq(-30,60),
                       launch_speed = 100) %>% 
  mutate(hit.prob = predict(fit, type = "response",
                 newdata = .))
batted100mph %>% 
  ggplot(aes(x = launch_angle,
             y = hit.prob)) +
  geom_line()
```


Here are hit probabilities for each batted ball in the 2017.

```{r, cache = TRUE, dev = 'png'}
batted2017 %>% 
  mutate(hit.prob = predict(fit, 
                            type = "response",
                            newdata = .)) -> batted2017

ev_plot2 <- batted2017 %>% 
  sample_n(nrow(.)/2) %>% 
  ggplot(aes(x = launch_speed, 
             y = launch_angle,
             color = hit.prob)) +
  geom_point() +
  scale_color_viridis() +
  geom_hline(data = guidelines,
             aes(yintercept = launch_angle),
             color = "black",
             linetype = 2) +
  geom_text(data = guidelines,
            aes(label = label, x = 0, y = launch_angle - 4),
            color = "black", hjust = "left")

ev_plot2
```

Let's find the batted ball with the highest hit probability that was not a hit.

```{r}
batted2017 %>% 
  filter(hit_flag == 0) %>% 
  arrange(-hit.prob) %>% 
  head(1) %>% 
  select(player_name, launch_angle, launch_speed, events, hit.prob, home_team)
```

