---
title: "Lesson_20_Boardsheet"
author: "Kevin Cummiskey"
date: "3/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Review

Last lesson, we discussed the difference between \textit{fixed effects} and \textit{random effects} in modeling catcher framing ability. 

Briefly explain the difference between a fixed effect and a random effect. \vspace{0.25in}

If you have 50 catchers, how many parameters are there in a fixed effects model? \vspace{0.25in}

If you have 50 catchers, how many parameteris are there in a random effects model? \vspace{0.25in}

If you wanted to know if Sandy Leon of the Boston Red Sox had a higher called strike percentage than other catchers, should you use a fixed or random effect? \vspace{0.25in}

If you wanted to know if the variability in catcher called strike percentage is ``interesting", should you use a fixed or random effect? \vspace{0.25in}

## Models with multiple random effects

Last class, we discussed a model for called strike percentage with a catcher random effect that adjusts for pitch location.

\[Strike_i \sim \text{Bernoulli}(p_i)\]
\[\log\left(\frac{p}{1-p}\right) = \beta_0 + f(px, pz) + catcher_j\]
\[catcher_j \sim \text{Normal}(0,\sigma^2)\]

### Read in the data

Follow the steps in Lesson 19 to read in Statcast data.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyverse)
library(devtools)

#Do this once to install baseballr package
#devtools::install_github("BillPetti/baseballr")
library(baseballr)

#read in data
pitches <- scrape_statcast_savant(start_date = "2017-05-01",
                                  end_date = "2017-05-10")
#called strikes and balls only
pitches %>% filter(type %in% c("S","B")) -> taken

#add catcher's name to the taken data.frame
path = "https://raw.githubusercontent.com/beanumber/baseball_R/master/data/"
file = "masterid.csv"
master_id <- read_csv(paste(path,file, sep = ""))

#merge with taken, note catcher's ID is the fielder_2_1 variable
taken %>% 
  left_join(select(master_id,mlb_id,mlb_name),
            by = c("fielder_2_1" = "mlb_id")) %>% 
  rename(catcher = mlb_name) -> taken
```

### Fit the model adjusting for pitch location

```{r, message = FALSE, fig.height= 2}
library(mgcv)
library(broom)
library(lme4)
library(ggrepel)
#fit pitch location model (plate_x = px, plate_z = pz)
model.location <- gam(type == "S" ~ s(plate_x, plate_z),
                      family = "binomial",
                      data = taken)
#get predictions from location model
taken %>% 
  mutate(strike_prob = predict(model.location, newdata = .,
                               type = "response")) -> taken

#fit random effects model adjusting for pitch location
model.random.adj <- glmer(type == "S" ~ strike_prob + (1|catcher),
                          family = "binomial",
                          data = taken)
# random effects
model.random.adj %>% tidy(effects = "ran_pars")
# fixed effects
model.random.adj %>% tidy(effects = "fixed")

#get random effects
model.random.adj %>%
  ranef() %>% 
  as_tibble() %>% 
  transmute(id = levels(grp),
            effect = condval) %>% 
  arrange(desc(effect)) -> catcher_effects.adj

catcher_effects.adj %>% head(4)
catcher_effects.adj %>% tail(4)

catcher_effects.adj %>% 
  ggplot(aes(x = effect, label = id)) +
  geom_dotplot() 
```

### Models with multiple random effects

What other variables should we adjust for?

\vspace{0.25in}

Write a model adjusting pitch location (fixed effect) and for pitcher using a random effect.

\vspace{0.25in}

```{r, fig.height=2}
#fit random effects model adjusting for pitch location and pitcher
model.random.adj2 <- glmer(type == "S" ~ strike_prob + (1|catcher) + (1|pitcher),
                          family = "binomial",
                          data = taken)
# random effects
model.random.adj2 %>% tidy(effects = "ran_pars")

# fixed effects
model.random.adj2 %>% tidy(effects = "fixed")

#get random effects
model.random.adj2 %>%
  ranef() %>% 
  as_tibble() %>% 
  filter(grpvar == "catcher") %>% 
  transmute(id = grp,
            effect = condval) %>% 
  arrange(desc(effect)) -> catcher_effects.adj2

catcher_effects.adj2 %>% head(4)
catcher_effects.adj2 %>% tail(4)

catcher_effects.adj2 %>% 
  ggplot(aes(x = effect, label = id)) +
  geom_dotplot() 
```

