---
title: "Lesson_19_Boardsheet"
author: "Kevin Cummiskey"
date: "3/3/2020"
output: pdf_document
---

## Does catchers' framing of pitches affect umpires' ball and strike calls?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read in Statcast data using the baseballr package

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(devtools)

#Do this once to install baseballr package
#devtools::install_github("BillPetti/baseballr")

# load baseballr package
library(baseballr)

#read in data
pitches <- scrape_statcast_savant(start_date = "2017-05-01",
                                  end_date = "2017-05-10")
#called strikes and balls only
pitches %>% filter(type %in% c("S","B")) -> taken

#add catcher's name to the taken data.frame
#get master ID list
path = "https://raw.githubusercontent.com/beanumber/baseball_R/master/data/"
file = "masterid.csv"
master_id <- read_csv(paste(path,file, sep = ""))

#merge with taken, note catcher's ID is the fielder_2_1 variable
taken %>% 
  left_join(select(master_id,mlb_id,mlb_name),
            by = c("fielder_2_1" = "mlb_id")) %>% 
  rename(catcher = mlb_name) -> taken
```

### How much variability is there in called strike probability by catcher?

First, let's see how much variability there is in called strike probabilities by catcher.  

```{r}
library(ggrepel)

# count balls and strikes by catcher
taken %>% 
  group_by(catcher) %>% 
  count(type) %>% 
  pivot_wider(id_cols = catcher, names_from = type, values_from = n) -> catchers

min.pitches = 300
# filter catchers with less than min.pitches
catchers %>% filter(S + B >= min.pitches) -> catchers

# calculate called strike probability and odds
catchers %>% 
  mutate(strike.prob = S/(S+B)) %>% 
  arrange(desc(strike.prob))-> catchers
catchers %>% head(5)
catchers %>% tail(5)
```

Plot the distribution.

```{r, fig.height = 2, message = FALSE}
catchers %>% 
  ggplot(aes(x = strike.prob)) +
  geom_dotplot() +
  labs(x = "strike probability",
       title = "Distribution of called strike probabilities by catcher") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

What do these results suggest?

\vspace{1in}

### Effect of catcher - Fixed effects

Thus far in your education, when dealing with regression, you've probably only encountered ``fixed effects".  Here is a fixed effects model we might fit to this data.

\[Strike_i \sim \text{Bernoulli}(p_i)\]
\[\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 catcher_1 + \beta_2 catcher_2 + \cdots + \beta_{m-1} catcher_{m-1}\]

where $catcher_j$ is an indicator (1/0) of whether catcher $j$ is the catcher on pitch $i$.

How many parameters are in this model? \vspace{0.1in}

What kinds of questions can we answer with this fixed effects model?

\vspace{0.5in}

Let's fit the fixed effects model.

```{r}
model.fixed <- glm(type == "S" ~ as.factor(catcher), 
                   family = "binomial", 
                   data = taken)
summary(model.fixed)
```

What do we conclude from these results?

\vspace{0.5in}

### Effect of catcher - random effects

The fixed effect model doesn't directly answer the question we are interested in.  Instead, we can say the catcher effect is itself a random variable.  

\[Strike_i \sim \text{Bernoulli}(p_i)\]
\[\log\left(\frac{p}{1-p}\right) = \beta_0 + catcher_j\]
\[catcher_j \sim \text{Normal}(0,\sigma^2)\]

where $catcher_j$ is the random effect of catcher $j$.

How many parameters are in this model? \vspace{0.1in}

What kinds of questions can we answer with this model? 

\vspace{0.5in}

Which parameter answers these questions? \vspace{0.1in}

Let's fit the random effects model.

```{r, message = FALSE, warning = FALSE}
library(lme4)

model.random <- glmer(type == "S" ~ (1|catcher), 
                      family = "binomial",
                      data = taken)
summary(model.random)
```

What are the estimates of the model parameters? 

\vspace{0.5in}

What do we conclude from this analysis? 

\vspace{0.5in}

Let's look at the catcher random effects.

```{r, fig.height = 2, warning = FALSE}
model.random %>%
  ranef() %>% 
  as_tibble() %>% 
  transmute(id = levels(grp),
            effect = condval) %>% 
  arrange(desc(effect)) -> catcher_effects
catcher_effects %>% head(5)
catcher_effects %>% tail(5)
catcher_effects %>% 
  ggplot(aes(x = effect)) +
  geom_dotplot()
```

### Confounding

OK, so far we've concluded the catcher makes a difference in called strike probability.  Let's say we compare a catcher with a called strike probability of 0.5 to another catcher with 0.6.  Would you conclude the difference is evidence one catcher is better at framing than the other? Explain.

\vspace{1in}

List variables would you adjust for to make better conclusions about catcher framing.

\vspace{0.5in}

### Adjusting for pitch location.

Write a model for called strike probability adjusting for strike probability with a catcher random effect.

\vspace{1in}

Instead of fitting the model above directly, we will first fit the pitch location model and then use predictions from it in the random effects model. 

```{r, message = FALSE, fig.height= 2}
library(mgcv)
library(broom)
#fit pitch location model (plate_x = px, plate_z = pz)
model.location <- gam(type == "S" ~ s(plate_x, plate_z),
                      family = "binomial",
                      data = taken)
#get predictions from location model
taken %>% 
  mutate(strike_prob = predict(model.location, newdata = .,
                               type = "response")) -> taken

#fit random effects model adjusting for pitch location
model.random.adj <- glmer(type == "S" ~ strike_prob + (1|catcher),
                          family = "binomial",
                          data = taken)
summary(model.random.adj)

#get random effects
model.random.adj %>%
  ranef() %>% 
  as_tibble() %>% 
  transmute(id = levels(grp),
            effect = condval) %>% 
  arrange(desc(effect)) -> catcher_effects.adj
catcher_effects.adj %>% head(5)
catcher_effects.adj %>% tail(5)
catcher_effects.adj %>% 
  ggplot(aes(x = effect)) +
  geom_dotplot() 
```
