---
title: "Lesson_12_Boardsheet"
author: "Kevin Cummiskey"
date: "2/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Review

Last class, we discussed the Run Expectancy Matrix and how to obtain it from the Retrosheet play-by-play data. Here is the Run Expectancy Matrix from last class.

```{r, message = FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(Lahman)

RUNS <- read_csv(file = "./data/expectedRUNs2016.csv")

RUNS_out = matrix(round(RUNS$Mean,2), 8,3)
colnames(RUNS_out) = c("0 outs", "1 out", "2 outs")
rownames(RUNS_out) = c("000","001","010","011",
                       "100","101","110", "111")
kable(RUNS_out)
```

Let's say there is a runner on 2nd base.  Stealing 3rd base is the most valuable when there are how many outs?

\vspace{0.5in}

## Measuring the Success of a Batting Play

We can estimate the value of a plate appearance as the difference in run expectancies between the new and old states plus the number of runs scored on the play. 

\[\text{RUN VALUE} = \text{RUNS}_\text{new state} -  \text{RUNS}_\text{old state} + \text{RUNS}_\text{scored on play}\]

Calculate the RUN VALUE of the following plays:

(1) There are runnners on first and second with no outs.  The batter successfully sacrifice bunts, resulting in runners on second and third with 1 out.    

&nbsp;

(2) There is a runner on second base with no outs.  The batter hits a single that scores the runner from second.  The batter stops at first base.

&nbsp;

(2) There is a runner on second base with two outs.  The batter hits a single that scores the runner from second.  The batter stops at first base.

&nbsp;

(2) What is the most valuable play in baseball? Explain.

Here's how to calculate the run value suing the code from last class (pg 116).

```{r, message = FALSE}
# load data2016 (as of page 115 in text) 
data2016 <- read_csv(file = "./data/data2016.csv")

# add run value of the play (page 116)
data2016 %>% 
  left_join(select(RUNS, -Outs), by = "STATE") %>%
  rename(Runs.State = Mean) %>% 
  left_join(select(RUNS, -Outs),
            by = c("NEW.STATE" = "STATE")) %>% 
  rename(Runs.New.State = Mean) %>% 
  replace_na(list(Runs.New.State = 0)) %>% 
  mutate(run_value = Runs.New.State - Runs.State + 
           RUNS.SCORED) -> data2016
```

Let's look at Jose Altuve (page 117).

```{r}
#get Altuve's playerID
Master %>%
  filter(nameFirst == "Jose", nameLast == "Altuve") %>% 
  pull(retroID) -> altuve.id

data2016 %>% 
  filter(BAT_ID == altuve.id,
         BAT_EVENT_FL == TRUE) -> altuve

altuve %>% 
  select(STATE, NEW.STATE, run_value) %>%
  slice(1:3)
```

Here's a plot of Altuve's run value by runner state.

```{r, fig.height = 3}
altuve %>%
  ggplot(aes(x = BASES, y = run_value)) +
  geom_jitter(width = 0.25, alpha = 0.5) +
  geom_hline(yintercept = 0, color = "blue") +
  xlab("RUNNERS")
```

Next, let's look at total runs by runners.

```{r}
altuve %>%
  group_by(BASES) %>% 
  summarize(RUNS = sum(run_value),
            PA = n()) -> Runs_Altuve
Runs_Altuve
```

Lastly, we can calculate a statistic called RE24 which sums RUNS over the different base runner combinations.

```{r}
Runs_Altuve %>% summarize(RE24 = sum(RUNS))
```

How do we interpret RE24?

\vspace{0.25in}

In baseball, there are context-neutral and context-dependent statistics.  Which is RE24?



