---
title: "Lesson_13_Boardsheet"
author: "Kevin Cummiskey"
date: "2/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Review

Last class, we discussed the run value of a play.  Write the equation for the run value of a play.

&nbsp;

Describe how you would calculate the RE24 for David Ortiz in 2016. Also, interpret this value in a way that's understandable to a general audience.

&nbsp;

Is it appropriate to compare David Ortiz's and Jose Altuve's RE24? Explain.

&nbsp;

## Value of a home run

Let's look at the value of a home run from a runs perspective.  First, here is the distribution of home runs by the number of runs scored.

```{r, message = FALSE, warning=FALSE}
library(tidyverse)
library(Lahman)

#recall data2016 contains a line for every play
data2016 <- read_csv("data/data2016.csv")

#get only home runs
data2016 %>% 
  filter(EVENT_CD == 23) -> home_runs

#number and percent of homeruns by runs scored
home_runs %>% 
  group_by(RUNS.SCORED) %>% 
  summarise(n = n()) %>%
  mutate(perc = round(n/sum(n),2))

#average number of runs scored
home_runs %>% 
  summarise(mean(RUNS.SCORED))
```

Explain why we might want to use the run value approach of Chapter 5 instead of the average runs scored approach above.  

\vspace{1in}

```{r}
#calculate the average run value of a home run
mean_hr <- home_runs %>% 
  summarise(mean_run_value = mean(run_value))
mean_hr
```

Why do the two approaches result in different answers?

\vspace{1in}

```{r, fig.height=2, message = FALSE}
home_runs %>% 
  ggplot(aes(run_value)) +
  geom_histogram() +
  geom_vline(data = mean_hr, aes(xintercept = mean_run_value),
             color = "blue", size = 1.5) +
  annotate("text", 1.7,2000, label = "Mean Run\nValue",
           color = "blue")
```
