---
title: "Lesson 24 Boardsheet - Multilevel Modeling"
author: "Kevin Cummiskey"
date: "3/25/2020"
output: 
    pdf_document:
urlcolor: blue
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today, we are going to investigate the limitations of the quadratic trajectories and discuss improvements using Bayesian statistics.  The reference for the next two lessons is:

* ``Multilevel Modeling of OBP trajectories'' by Jim Albert. https://baseballwithr.wordpress.com/2019/11/25/multilevel-modeling-of-obp-trajectories/

Learning Objectives:

* Gain appreciation for how Bayesian statistics can help us combine prior knowledge with new observations to update our beliefs.

* Gain appreciation for how multilevel modeling can ``pool'' information to arrive a better estimates for individuals.

## Players who debuted in 2001.

Let's find the players who debuted in the year 2001.

```{r, message= FALSE, warning = FALSE} 
library(Lahman)
library(tidyverse)
library(lubridate)
library(ggrepel)

#Players with debut in year 2000
Master %>% 
  filter(year(debut) == 2001) %>% 
  pull(playerID) -> year2001.ids

#Players with at least 1000 atbats
Batting %>% 
  filter(playerID %in% year2001.ids) %>% 
  group_by(playerID) %>% 
  summarize(AB = sum(AB)) %>% 
  filter(AB > 1000) %>% 
  pull(playerID) -> player.ids
```

The players who debuted in 2001 who would go on to have at least 1000 at bats are:

```{r}
library(knitr)
Master %>% 
  filter(playerID %in% player.ids) %>% 
  select(nameFirst, nameLast) %>% 
  kable()
```

## OBP trajectories

Let's look at their OBP trajectories using the quadratic model fit individually to players.

```{r, fig.height= 3}
source("Chapter8_functions.R")

# get statistics by age and add names
player.ids %>% 
  map_df(get_stats) %>% 
  left_join(Master %>% select(nameLast, nameFirst, playerID))-> player.stats

#plot trajectories
player.stats %>% 
  ggplot(aes(x = Age, y = OBP, group = playerID)) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              se = FALSE)
```

What do you think of these models?

\vspace{1in}


How do we fix these issues?

\newpage

## Bayesian Approach

\newpage

## Trajectories from Multilevel Models

Now, let's fit trajectories that pool information from all the players in the data set.

```{r, cache = TRUE}
library(brms)
library(rstan)
player.stats %>% 
  mutate(AgeD = Age - 30,
         Player = paste(nameFirst,nameLast, sep = " ")) -> player.stats


fit <- brm(OB | trials(PA) ~ AgeD + I(AgeD ^ 2) + 
             (AgeD + I(AgeD ^ 2) | Player),
           data = player.stats,
           family = binomial("logit"))

Player_Fits <- coef(fit)$Player[, "Estimate", ] %>% 
  as_tibble(rownames = "Player") %>% 
  rename(b0.hat = Intercept,
         b1.hat = AgeD,
         b2.hat = IAgeDE2)

# merge these estimates with our main dataset

player.stats <- inner_join(player.stats, Player_Fits, by = "Player")

# find estimates of OBP probs at each age
# note plogis is the logit function
player.stats %>% 
  mutate(OBP.pred = plogis(b0.hat + b1.hat * AgeD + b2.hat * AgeD^2)) -> player.stats

player.stats %>% 
  ggplot(aes(x = Age,
             y = OBP.pred,
             group = Player)) +
  geom_line(color = "red")
```

What do you think of the new trajectories?

\vspace{1in}


Let's focus on one player who had a really weird individual trajectory:

```{r}
player.stats %>%
  split(pull(.,playerID)) %>% 
  map(fit_model)

player.stats %>% 
  filter(playerID == "custja01") %>%
  ggplot(aes(x = Age, y = OBP)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2), se = FALSE) +
  ylim(0,1) + xlim(20,40) +
  geom_line(aes(y = OBP.pred), color = "red")
```


```{r, eval = FALSE, include = FALSE}
# find peak ages using two models

player.stats %>% group_by(Player) %>% 
  summarize(B0 = first(B0),
            B1 = first(B1),
            B2 = first(B2),
            b0 = first(Intercept),
            b1 = first(AgeD.y),
            b2 = first(IAgeDE2)) %>% 
  mutate(Ind_Peak_Age = 30 - B1 / 2 / B2,
           MLM_Peak_Age = 30 - b1 / 2 / b2)  %>% 
  select(Ind_Peak_Age, MLM_Peak_Age) %>% 
  pivot_longer(everything(),
               names_to = "Type",
               values_to = "Age") -> S

# graph two sets of peak ages

ggplot(S, aes(Type, Age)) +
  geom_point() +
  coord_flip() +
  increasefont() +
  ggtitle("Estimates of Peak Age") +
  centertitle()

```


