---
title: "Lesson 3 - Intro to R tidyverse"
author: "Kevin Cummiskey"
date: "January 14, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Lahman)
```

## Review

Last class, we discussed the five tidyverse verbs (filter, select, mutate, summarize, arrange).  Use the Pitching database in the Lahman package to generate the table (Pitchers with 350 wins or more) on page 65 of the Marchi text.  Note the Pitching database contains one row for each pitcher for each season or stint (if the pitcher played for more than one team in a season).  For this problem, you can use playerID instead of full name in the table.  Write your code below.

\vspace{2in}

```{r, include = FALSE}
Pitching %>%
  group_by(playerID) %>%
  summarize(W = sum(W, na.rm = TRUE),
            L = sum(L, na.rm = TRUE),
            SO = sum(SO, na.rm = TRUE),
            BB = sum(BB, na.rm = TRUE)) %>%
  filter(W > 349) -> Wins.350
```

How could you add the player's full name?  (hint: use the Master database.)

\vspace{1in}


```{r, include = FALSE}
Wins.350 %>% 
  left_join(Master %>% select(playerID, nameFirst, nameLast),
            by = "playerID") %>%
  mutate(player = paste(nameFirst,nameLast)) %>%
  select(player, W, L, SO, BB) -> Wins.350
```

## Pitching

There has been a lot of discussion recently about the evolving use of relief pitchers.  Historically, starting pitchers were expected to go deep into the game and they frequently pitched the whole game.  In the last several years, teams have begun employing relief pitchers in highly specialized roles.  For a nice discussion on this, see Nate Silver's article ``Relievers Have Broken Baseball. We Have A Plan To Fix It." (\url{https://fivethirtyeight.com/features/relievers-have-broken-baseball-we-have-a-plan-to-fix-it/}) 

1. What are some questions we would want to investigate to better understand how the use of relief pitchers has changed over time?

\vspace{2in}

### Complete Games

For each season from 1970 to 2018, calculate the fraction of games each year that were complete games (hint: Teams database). Discuss two interesting results.

\vspace{1in}

```{r, include = FALSE}
Teams %>%
  filter(yearID > 1969) %>%
  group_by(yearID) %>%
  summarize(CG.frac = sum(CG)/sum(G)) -> complete.games

complete.games %>%
  ggplot(aes(x = yearID,y = CG.frac)) +
  geom_point()
```

Does the trend differ between the American and National Leagues? 

\vspace{0.5in}

```{r, include=FALSE}
Teams %>%
  filter(yearID > 1969) %>%
  group_by(yearID,lgID) %>%
  summarize(CG.frac = sum(CG)/sum(G),) -> complete.games.lg

complete.games.lg %>%
  ggplot(aes(x = yearID,
             y = CG.frac,
             color = lgID)) +
  geom_point()
```

### Batters Faced

```{r}
Pitching %>%
  filter(yearID > 1969) %>%
  group_by()
```



