---
title: "Required Exercises 1"
author: "Jason Agsalud"
date: "1/26/2020"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lesson 1 Question 1
1a. In order to find how the rate of walks has changed over the history of baseball, one would use the Lahman database. Specifically, by using the "Teams" sub-data base, one would be able to analyze this change. 

1b.  In order to determine the number of shutouts thrown during the 1968 season, one should use the Retrosheet game logs. By doing so, you would be able to extrapolate specific in-game statistics and then filter between the 2012 and 1968 season to compare shutouts thrown in those specific years. 

1c. Because of the specificity of this question, one should not use the Lahman database, but rather use the PITCHf/x database.First, we would need to filter to only show data relevant to the first pitches of each at bat. Then, we could apply the filter to only first-pitch strikes. From this, we can determine the ratio of first-pitch strikes. Furthermore, to determine tthe fraction of the pitches that are strikes after a 2-0 account, we would need to show the next pitch after two consectuve pitches out of the strike zone. 

1d. The Statcast database would be best to analyze which players are most likely to hit grownballs. First, you could filter by position type and compare the different positions. Next you would then arrange the data of the launch speed of the hits and put them in descending order. 

1e. By using the Lahman database, you can find the number of stolen bases to second base. Then, you would need to find the number of stolen bases to third bases. From here, you can then mutate the data and create a new variable to determine the ratio and see which base is easier to steal.

## Lesson 1 Question 4
The first batter "vansa001" took a first pitch strike then watched three straight balls. On a 3-1 count, he hit a line drive to left field and recorded a single. The next batter "bondb001" took a first pitch strike too, then swung and missed making the count 0-2. He watched one ball before putting the ball in play in the 6- hole between the shortstop and the third baseman. With runners now on first and second, the final batter worked a full count before a triple play was recorded to end the inning. 

## Lesson 1 Question 5
The knuckleball is a unique pitch in baseball because it is the only pitch that has little to no spin when it is delivered from the pitcher to the catcher. As a result, a unorthodox and sometimes unpredictable flight pattern occurs. A knuckleball can go up and down, side to side, or sometimes a combination of either, making it difficult for the cather to predict where to catach the ball as well as the hitter predicting when and where to swing. A fastball on the other hand has backspin and usually tracks in a straight line. From this data, you can see a drastic drop-off in speed where a normal MLB fastball is around 90 mph, and R.A. Dickey's fastball ranged from 73-81 mph. Additionally, we can see that the vertical and horizontal change in plane of the ball path ventured severely away from 0.  


## Lesson 2 Question 1
```{r}
SB <- c(1406, 938, 897, 741, 738, 689, 506, 504, 474)
CS <- c(335, 307, 212, 195, 109, 162, 136, 131, 114)
G <- c(3081, 2616, 3034, 2826, 2476, 2649, 2599, 2683, 2379)

SB.Attempt <- SB + CS

Success.Rate <- SB / SB.Attempt

SB.Game <- SB / G

plot(SB.Game, Success.Rate)
```

Max Carey had the highest stolen base success rate with a rate of stealing a base nearly 87% of the time. The player with the highest number of stolen bases per game was Ricky Henderson (.45 steals per game).

## Lesson 2 Question 2 
```{r}
outcomes <- c("single", "out", "out", "single", "double", "out", "walk", "out", "single")
table(outcomes)
```

```{r}
f.outcomes <- factor(outcomes, levels = c("out", "walk", "single", "double"))
table(f.outcomes)
```

The difference in the two tables is the order in which the events appear in the list. The second table was arranged in a more ordered manner in respect to total base where an "out" receives 0 total bases, "walk" and "single" is one total base, and a "double" is two total bases. 

```{r}
outcomes == "walk"
sum(outcomes == "walk")
```

By altering the variable "outcomes" to the above line of code, we see that we are limiting the outcomes of an at-bat to just only walks, and then we are summing up the number of times a batter walks. 

## Lesson 3 Question 3
```{r}
W <- c(373, 354, 364, 417, 355, 373, 361, 363, 511)
L <- c(208, 184, 310, 279, 227, 188, 208, 245, 316)
Name <- c("Alexander", "Clemens", "Galvin", "Johnson", "Maddux", "Mathewson", "Nichols", "Spahn", "Young")

Win.PCT <- 100*(W/(W+L))
Wins.350 <- data.frame(Name, W, L, Win.PCT)
```

```{r}
Wins.350 %>%
  arrange(desc(Win.PCT))
```
From this result, we can see that Christy Matthewson had the highest Win Percentage and Pud Galvin had the lowest Win Percentage amongst this list of pitchers. 

## Lesson 3 Question 4
```{r}
SO <- c(2198, 4672, 1806, 3509, 3371, 2502, 1868, 2583, 2803)
BB <- c(951, 1580, 745, 1363, 999, 844, 1268, 1434, 1217)

SO.BB.Ratio <- SO/BB

SO.BB <- data.frame(Name, SO, BB, SO.BB.Ratio)

SO.BB %>%
  filter(SO.BB.Ratio > 2.8) %>%
  arrange(desc(BB))
```

Interestingly, when you arrange the data frame by number of walks, it is easier to see that the pitcher with the lowest Strikeout to Walk ratio also had a high number of walks thrown and vice versa.

##Lesson 3 Question 5
```{r}
career.pitching <- Pitching %>%
  group_by(playerID) %>%
  summarize(SO = sum(SO, na.rm = TRUE),
            BB = sum(BB, na.rm = TRUE),
            IPouts = sum(IPouts, na.rm = TRUE),
            midYear = median(yearID, na.rm = TRUE))
```

```{r}
install.packages("tidyverse")
library(tidyverse)
```